Neural Circuit Organization of the Ctenophore Apical Organ and Its Role in Gravity Sensing

### 2025-03-16 manuscript/Manuscript.qmd
edit ref

Among these, we focused on the apical organ of ctenophores. In many bilaterian and non-bilaterian animals, an apical nervous system is located at one pole of the body, forming a specialized sensory structure that governs physiological functions and motor activities. In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974] <!--# (Tamm, 1982 ?) -->. Additionally, immunostaining using an anti-tyrosylated α-tubulin antibody has identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].

In this study, we specifically examined the gravity-sensing system within the apical organ. The apical organ of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014]. The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014] <!--# (Tamm, 1982 ?) -->. Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron]. Therefore, we conducted a connectome analysis to characterize the network structure of the apical organ and elucidate the neural circuits that link sensory input to behavioral responses in ctenophores.

@@

As mentioned above, following the identification of the presynaptic structures, we reconstructed three major Syncytial neurons. Each of these Syncytial neurons was a cell with multiple nuclei, with membranes fused by continuous plasma membranes. These neurons are distinct in morphology from the Syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously reconstructed in 3D [@Burkhardt_2023]. Our findings represent the second documented discovery of Syncytial neurons in ctenophores. Furthermore, these neurons exhibited clear morphological differences when compared to other sensory cells reported in the same study, such as mesogleal neurons and sensory cells with presynaptic structures and cilia (types 1-4)[@Burkhardt_2023]. Based on their distinct spatial relationships, we were able to classify these three Syncytial neurons into two categories. The first type is a larger "AO neuron_Q1234," which possesses four (or possibly six?) nuclei spanning four quadrants. It contained X presynaptic structures. The second type consists of "AO neuron_Q12" and "AO neuron_Q34," each having two nuclei spanning two quadrants, and each containing X presynaptic structures.

@@

To investigate the function of gravity-sensitive circuits, we conducted a comparative analysis of balancer ciliary movements across different regions. Previous studies (Tamm, 1980, 1982; Lowe, 1997) have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Furthermore, it has been suggested that differences in the asymmetry of statolith morphology and the elongation direction of balancer cilia between the tentacular and sagittal planes could result in variations in the force received from the statolith (Tamm, 2014, 2015). Building on these findings, we standardized conditions for observing balancer cilia by tilting the microscope by 90 degrees and fixing samples such that the aboral-oral axis was aligned parallel to the vertical stage, with the aboral side facing upward. Using connectome analysis, we aimed to evaluate the neural circuit differences by analyzing the control patterns of left and right balancer cilia in the tentacular and sagittal planes. Specifically, we hypothesized that differences in control between Q1 (Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3) regions, would reveal distinct functions of AO neurons (e.g., AO neuron Q12/Q34 vs. AO neuron Q1234). For the sagittal plane, we recorded balancer ciliary movements in the Q1(4) and Q2(3) regions (left and right sides of the image) and, for the tentacular plane, in the Q1(3) and Q2(4) regions, using a high-speed camera (100 fps) for 2 minutes (12,000 frames). Regions of interest (ROIs) were defined in areas where brightness changes indicated ciliary beating. We measured these brightness changes to generate dynamics graphs of balancer ciliary movements for both planes.

@@

We identified three syncytial neurons with multiple nuclei in the apical organ of five-day-old *M. leidyi* cydippid larvae. Burkhurdt et al. (2023) previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia. However, the newly discovered syncytial neurons embedded within the apical organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons. Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa (Mayorova et al., 2021), with cell bodies extending through intercellular spaces. These morphological characteristics suggest that the newly identified neurons represent a second distinct type of syncytial neuron in ctenophores.

@@

The three identified neurons may be classified into two distinct functional types based on their spatial arrangement and their influence on the differential control of balancer ciliary movement between the quadrants. In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements. For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating (Verasztó et al., 2017). Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells. Activation of Ser-h1 induces ciliary beating initiation and increases beat frequency.

@@

Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e] <!--#  Krisch, 1973; Aronova, 1974 ? --> ). By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the apical organ. However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets. In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ [@Sachkova2021; @Hayakawa2022]. These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis (Jekely, 2021). While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.

Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons [@Burkhardt2023]. Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells <!--# add these to bib file -->(Hertwig, 1880; Chun, 1880; Eimer, 1880).

@@

The neural cells identified in the apical organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets. Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution (Luo, 2021). Moreover, the presence of multiple divergent feedforward circuits suggests that slight variations in the synchronization and phase delay of ciliary beating (within the range of a few hundred milliseconds) may have evolved as a mechanism for generating behavioral diversity.

From a broader neuroanatomical perspective, nervous system structures across various animals have been shaped through selective optimization to adapt to environmental constraints (Bullmore & Sporns, 2012; Perin et al., 2011). For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring (White et al., 1986). Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains [@Farnworth2024]. In particular, this research may help elucidate how circuit modularization and layered structures emerged over evolutionary time.

@@

The reconstruction of cells and connectivity was performed using CATMAID (Saalfeld et al., 2009), analyzing a total of 620 serial sections. To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server [@Saalfeld2009; @Schneider-Mizell2016]. To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset. At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer. All skeletons were rooted at their respective cell bodies, and nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria. Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.

>>>>

Among these, we focused on the apical organ of ctenophores. In many bilaterian and non-bilaterian animals, an apical nervous system is located at one pole of the body, forming a specialized sensory structure that governs physiological functions and motor activities. In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974; @Tamm1982]. Additionally, immunostaining using an anti-tyrosylated α-tubulin antibody has identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].

In this study, we specifically examined the gravity-sensing system within the apical organ. The apical organ of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014]. The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014; @Tamm1982]. Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron]. Therefore, we conducted a connectome analysis to characterize the network structure of the apical organ and elucidate the neural circuits that link sensory input to behavioral responses in ctenophores.

@@

As mentioned above, following the identification of the presynaptic structures, we reconstructed three major Syncytial neurons. Each of these Syncytial neurons was a cell with multiple nuclei, with membranes fused by continuous plasma membranes. These neurons are distinct in morphology from the Syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously reconstructed in 3D [@Burkhardt_2023]. Our findings represent the second documented discovery of Syncytial neurons in ctenophores. Furthermore, these neurons exhibited clear morphological differences when compared to other sensory cells reported in the same study, such as mesogleal neurons and sensory cells with presynaptic structures and cilia (types 1-4)[@Burkhardt_2023]. Based on their distinct spatial relationships, we were able to classify these three Syncytial neurons into two categories. The first type is a larger "AO neuron_Q1234," which possesses four (or possibly five?) nuclei spanning four quadrants. It contained X presynaptic structures. The second type consists of "AO neuron_Q12" and "AO neuron_Q34," each having two nuclei spanning two quadrants, and each containing X presynaptic structures.

@@

To investigate the function of gravity-sensitive circuits, we conducted a comparative analysis of balancer ciliary movements across different regions. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Furthermore, it has been suggested that differences in the asymmetry of statolith morphology and the elongation direction of balancer cilia between the tentacular and sagittal planes could result in variations in the force received from the statolith [@Tamm_2014; @Tamm2015]. Building on these findings, we standardized conditions for observing balancer cilia by tilting the microscope by 90 degrees and fixing samples such that the aboral-oral axis was aligned parallel to the vertical stage, with the aboral side facing upward. Using connectome analysis, we aimed to evaluate the neural circuit differences by analyzing the control patterns of left and right balancer cilia in the tentacular and sagittal planes. Specifically, we hypothesized that differences in control between Q1 (Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3) regions, would reveal distinct functions of AO neurons (e.g., AO neuron Q12/Q34 vs. AO neuron Q1234). For the sagittal plane, we recorded balancer ciliary movements in the Q1(4) and Q2(3) regions (left and right sides of the image) and, for the tentacular plane, in the Q1(3) and Q2(4) regions, using a high-speed camera (100 fps) for 2 minutes (12,000 frames). Regions of interest (ROIs) were defined in areas where brightness changes indicated ciliary beating. We measured these brightness changes to generate dynamics graphs of balancer ciliary movements for both planes.

@@

We identified three syncytial neurons with multiple nuclei in the apical organ of five-day-old *M. leidyi* cydippid larvae. Burkhurdt et al. previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia [@Burkhardt2023]. However, the newly discovered syncytial neurons embedded within the apical organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons. Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa [@Mayorova2021], with cell bodies extending through intercellular spaces. These morphological characteristics suggest that the newly identified neurons represent a second distinct type of syncytial neuron in ctenophores.

@@

The three identified neurons may be classified into two distinct functional types based on their spatial arrangement and their influence on the differential control of balancer ciliary movement between the quadrants. In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements. For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating [@Verasztó2017]. Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells. Activation of Ser-h1 induces ciliary beating initiation and increases beat frequency.

@@

Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e; @Krisch1973; Aronova1974]. By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the apical organ. However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets. In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ [@Sachkova2021; @Hayakawa2022]. These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis [@Jékely2021]. While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.

Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons [@Burkhardt2023]. Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells [@Hertwig1880; @Chun1880; @Eimer1880].

@@

The neural cells identified in the apical organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets. Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution [@Luo2021]. Moreover, the presence of multiple divergent feedforward circuits suggests that slight variations in the synchronization and phase delay of ciliary beating (within the range of a few hundred milliseconds) may have evolved as a mechanism for generating behavioral diversity.

From a broader neuroanatomical perspective, nervous system structures across various animals have been shaped through selective optimization to adapt to environmental constraints [@Bullmore2012; @Perin2011]. For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring [@White1986]. Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains [@Farnworth2024]. In particular, this research may help elucidate how circuit modularization and layered structures emerged over evolutionary time.

@@

The reconstruction of cells and connectivity was performed using CATMAID [@Saalfeld2009], analyzing a total of 620 serial sections. To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server [@Saalfeld2009; @Schneider-Mizell2016]. To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset. At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer. All skeletons were rooted at their respective cell bodies, and nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria. Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.
